---
title: "FitBit Charge HR Accuracy During Exercise"
author: "Christopher Swingley"
date: '2016-01-09'
output:
  pdf_document:
    highlight: default
    includes:
      in_header: rmarkdown_tex_preamble.tex
    keep_tex: yes
    latex_engine: xelatex
    toc: no
  html_document:
    theme: united
    toc: yes
  word_document: default
---

# Introduction

This week a class action lawsuit was filed against FitBit, claiming that their
heart rate fitness trackers don’t perform as advertised, specifically during
exercise.  I’ve been wearing a FitBit Charge HR since October, and also wear a
Scosche Rhythm+ heart rate monitor whenever I exercise, so I have a lot of data
that I can use to assess the legitimacy of the lawsuit.

# The Data

Heart rate data from the Rhythm+ is collected by the RunKeeper app on my phone,
and after transferring the data to RunKeeper’s website, I can download GPX
files containing all the GPS and heart rate data for each exercise.  Data from
the Charge HR is a little harder to get, but with the proper tools and
permission from FitBit, you can get what they call “intraday” data.  I use the
`fitbit` Python library and a set of routines I wrote (available on
[GitHub](https://github.com/cswingle/fb2psql)) to pull this data.

The data includes 116 activities, mostly from commuting to and from work by
bicycle, fat bike, or skis.  The first step in the process is to pair the two
data sets, but since the exact moment when each sensor recorded data won’t
match, I grouped both sets of data into 15-second intervals, and calculated the
mean heart rate for each sensor withing that 15-second window.  The result
looks like this:

```{r echo=TRUE, eval=TRUE, warning=FALSE}
load("matched_hr_data.rdata")
kable(head(matched_hr_data))
```

Let’s take a quick look at a few of these activities. The squiggly lines show
the heart rate data from the two sensor, and the horizontal lines show the
average heart rate for the activity.  In both cases, the FitBit Charge HR is
shows in red and the Scosche Rhythm+ is blue.

```{r echo=TRUE, eval=TRUE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(scales)

heart_rate <- matched_hr_data %>%
    transmute(dt=dt_rounded, track_id=track_id,
              title=paste(strftime(as.Date(dt_rounded,
                                           tz="US/Alaska"), "%b-%d"),
                          type),
              fitbit=fitbit_hr, rhythm=rhythm_hr) %>%
    gather(key=sensor, value=hr, fitbit:rhythm) %>%
    filter(track_id %in% c(3587, 3459, 3437, 3503))

activity_means <- heart_rate %>%
    group_by(track_id, sensor) %>%
    summarize(hr=mean(hr))

facet_labels <- heart_rate %>% select(track_id, title) %>% distinct()
hr_labeller <- function(values) {
     lapply(values, FUN=function(x) (facet_labels %>% filter(track_id==x))$title)
}
r <- ggplot(data=heart_rate,
            aes(x=dt, y=hr, colour=sensor)) +
    geom_hline(data=activity_means, aes(yintercept=hr, colour=sensor), alpha=0.5) +
    geom_line() +
    theme_bw() +
    scale_color_brewer(name="Sensor",
                       breaks=c("fitbit", "rhythm"),
                       labels=c("FitBit Charge HR", "Scosche Rhythm+"),
                       palette="Set1") +
    scale_x_datetime(name="Time") +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
    scale_y_continuous(name="Heart rate (bpm)") +
    facet_wrap(~track_id, scales="free", labeller=hr_labeller, ncol=1) +
    ggtitle("Comparison between heart rate monitors during a single activity")

print(r)
```
```{r echo=FALSE, eval=TRUE, warning=FALSE}
height <- 9
width <- 16
scaling <- 0.75
svg("two_good_two_bad.svg", height=height*scaling, width=width*scaling)
print(r)
dev.off()
```

You can see that for each activity type shown here, one of the plots shows data
where the two heart rate monitors track well, and one where they don’t.  And
when they don’t agree, the FitBit is *wildly* inaccurate. When I initially got
my FitBit I experimented with different positions on my arm for the device, but
it didn’t seem to matter, so I settled on the advice from FitBit, which is to
place the band slightly higher on the wrist (two to three fingers from the
wrist bone) than in normal use.

One other pattern is evident from the two plots where the FitBit does poorly:
the heart rate readings are always much lower than reality.

A scatterplot of all the data, plotting the FitBit heart rate against the
Rhythm+ will show the overall pattern.

```{r echo=TRUE, eval=TRUE, warning=FALSE}
q <- ggplot(data=matched_hr_data,
            aes(x=rhythm_hr, y=fitbit_hr, colour=type)) +
    geom_abline(intercept=0, slope=1) +
    geom_point(alpha=0.25, size=1) +
    geom_smooth(method="lm", inherit.aes=FALSE, aes(x=rhythm_hr, y=fitbit_hr)) +
    theme_bw() +
    scale_x_continuous(name="Scosche Rhythm+ heart rate (bpm)") +
    scale_y_continuous(name="FitBit Charge HR heart rate (bpm)") +
    scale_colour_brewer(name="Activity type", palette="Set1") +
    ggtitle("Comparison between heart rate monitors during exercise")

print(q)
```
```{r echo=FALSE, eval=TRUE, warning=FALSE}
height <- 9
width <- 16
scaling <- 0.75
svg("fitbit_vs_rhythm_scatterplot.svg", height=height*scaling, width=width*scaling)
print(q)
dev.off()
```

If the FitBit device were always accurate, the points would all be distributed
along the 1:1 line, which is the diagonal black line under the point cloud.
The blue diagonal line shows the actual linear relationship between the FitBit
and Rhythm+ data.  What’s curious is that the two lines cross near 100 bpm,
which means that the FitBit is underestimating heart rate when my heart is
beating fast, but *over*estimating it when it’s not.

The color of the points indicate the type of activity for each point, and you
can see that most of the lower heart rate points (and overestimation by the
FitBit) come from hiking activities.  Is it the type of activity that triggers
over- or underestimation of heart rate from the FitBit, or is is just that all
the lower heart rate activities tend to be hiking?

Another way to look at the same data is to calculate the difference between the
Rhythm+ and FitBit and plot those anomalies against the actual (Rhythm+) heart
rate.

```{r echo=TRUE, eval=TRUE, warning=FALSE}
anomaly_by_hr <- matched_hr_data %>%
    mutate(anomaly=fitbit_hr-rhythm_hr) %>%
    select(rhythm_hr, anomaly, type)

q <- ggplot(data=anomaly_by_hr,
            aes(x=rhythm_hr, y=anomaly, colour=type)) +
    geom_abline(intercept=0, slope=0, alpha=0.5) +
    geom_point(alpha=0.25, size=1) +
    theme_bw() +
    scale_x_continuous(name="Scosche Rhythm+ heart rate (bpm)",
                       breaks=pretty_breaks(n=10)) +
    scale_y_continuous(name="Difference between FitBit Charge HR and Rhythm+ (bpm)",
                       breaks=pretty_breaks(n=10)) +
    scale_colour_brewer(palette="Set1")

print(q)
```
```{r echo=FALSE, eval=TRUE, warning=FALSE}
height <- 9
width <- 16
scaling <- 0.75
svg("fitbit_anomaly_vs_rhythm_scatterplot.svg", height=height*scaling, width=width*scaling)
print(q)
dev.off()
```

In this case, all the points should be distributed along the zero line (no
difference between FitBit and Rhythm+).  We can see a large bluish (fat biking)
cloud around the line between 130 and 165 bpm (indicating good results from the
FitBit), but the rest of the points appear to be well distributed along a
diagonal line which crosses the zero line around 90 bpm.  It’s another way of
saying the same thing: at lower heart rates the FitBit tends to overestimate
heart rate, and as my heart rate rises above 90 beats per minute, the FitBit
underestimates heart rate to a greater and greater extent.

# Student’s t-test and results

A Student’s t-test can be used effectively with paired data like this to judge
whether the two data sets are statistically different from one another.  This
routine runs a paired t-test on the data from each activity, testing the null
hypothesis that the FitBit heart rate values are the same as the Rhythm+
values.  I’m tacking on significance labels typical in analyses like these
where one asterisk indicates the results would only happen by chance 5% of the
time, two asterisks mean random data would only show this pattern 1% of the
time, and three asterisks mean there’s less than a 0.1% chance of this
happening by chance.

One note: There are 116 activities, so at the 0.05 significance level, we would
expect five or six of them to be different just by chance.  That doesn’t mean
our overall conclusions are suspect, but you do have to keep the number of
tests in mind when looking at the results.

```{r echo=TRUE, eval=TRUE, warning=FALSE}
t_tests <- matched_hr_data %>%
    group_by(track_id, type, min_temp, max_temp) %>%
    summarize_each(funs(p_value=t.test(., rhythm_hr, paired=TRUE)$p.value,
                        anomaly=t.test(., rhythm_hr, paired=TRUE)$estimate[1]),
                   vars=fitbit_hr) %>%
    ungroup() %>%
    mutate(sig=ifelse(p_value<0.001, '***',
                      ifelse(p_value<0.01, '**',
                             ifelse(p_value<0.05, '*', '')))) %>%
    select(track_id, type, min_temp, max_temp, anomaly, p_value, sig)
kable(head(t_tests))
```

It's easier to interpret the results summarized by activity type:

```{r echo=TRUE, eval=TRUE, warning=FALSE}
t_test_summary <- t_tests %>%
    mutate(different=grepl('\\*', sig)) %>%
    select(type, anomaly, different) %>%
    group_by(type, different) %>%
    summarize(n=n(),
              mean_anomaly=mean(anomaly))
kable(t_test_summary)
```

What this shows is that the FitBit underestimated heart rate by an average of
21 beats per minute in 26 of 28 (93%) bicycling trips, underestimated heart
rate by an average of 15 bpm in 58 of 73 (79%) fat biking trips, overestimate
heart rate by an average of 11 bpm in 80% of hiking trips, and always
drastically underestimated my heart rate while skiing.

For all the data:

```{r echo=TRUE, eval=TRUE, warning=FALSE}
t.test(matched_hr_data$fitbit_hr, matched_hr_data$rhythm_hr, paired=TRUE)
```

Indeed, in aggregate, the FitBit does a poor job at estimating heart rate
during exercise.

# Conclusion

Based on my data of more than 100 activities, I’d say the lawsuit has some
merit.  I only get accurate heart rate readings during exercise from my FitBit
Charge HR about 16% of the time, and the error in the heart rate estimates
appears to get worse as my actual heart rate increases.  The advertising for
these devices gives you the impression that they’re designed for high
intensity exercise (showing people being very active, running, bicycling,
etc.), but their performance during these activities is pretty poor.

All that said, I knew this going in when I bought my FitBit, so I’m not hugely
disappointed.  There are plenty of other benefits to monitoring the data from
these devices (including non-exercise heart rate), and it isn’t a major
inconvenience for me to strap on a more accurate heart rate monitor for those
times when it actually matters.

